---
title: "Early versus late tracheostomy study"
author: "Adrian Barnett"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: word_document
---

```{r setup, include=FALSE}
##
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE, comment='', dpi=400)
options(width=1000, scipen = 999) # Wide pages and no scientific numbers
Missing = function(x) base::sum(is.na(x))
Mean = function(x) base::mean(x, na.rm=TRUE)
Median = function(x) stats::quantile(x, probs=0.5, na.rm=TRUE)
Q1 = function(x) stats::quantile(x, probs=0.25, na.rm=TRUE)
Q3 = function(x) stats::quantile(x, probs=0.75, na.rm=TRUE)
Min = function(x) base::min(x, na.rm=TRUE)
Max = function(x) base::max(x, na.rm=TRUE)
Sum = function(x) base::sum(x, na.rm=TRUE)
SD = function(x) stats::sd(x, na.rm=TRUE)
N = function(x) base::length(x)
library(survival)
library(broom)
library(diagram)
library(etm) # to estimate change in time after tracheostomy
library(dplyr)
library(tables)
library(pander)
panderOptions('table.emphasize.rownames', FALSE)
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ',')
library(ggplot2)
cbPalette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
library(survminer) # for Kaplan-Meier
# function to run change in extra time (Beyersmann/Wolkewitz/Allignol)
source('1_extra_time.R')
# function to round with trailing zeros
roundz  = function(x, digits=0){formatC( round( x, digits ), format='f', digits=digits)}
## main data:
load('data/AnalysisReady.RData') # from 0_read_data.R
```

# Methods

## Event times

The key sequence of events is shown in the figure below.

```{r diagram, fig.width=5, fig.height=4}
source('1_event_diagram.R')
```

By "outcome" we mean a key outcome of interest, such as eating or starting physiotherapy. Some patients may go straight from being ventilated to the outcome, whereas others may first receive a tracheostomy. Our key research interest is whether the tracheostomy hastens the time to the outcome. We measure this by examining the days between these key events for all patients. The start of ventilation is the first event for every patient.

Here tracheostomy is known as an "intermediate event" because it occurs between the start event (ventilation) and the outcome of interest. To correctly estimate the change in time after tracheostomy we used a method that accounts for the varying time of tracheostomy. We used the multi-state modelling approach described in detail by Beyersmann et al (2012) and used the `etm' package in R to estimate the change in length of stay after tracheostomy (Allignoal et al,  2011).

We made the following assumptions concerning two events occurring on the same day:

```{r table.assumption}
f = data.frame(`Event 1` = c('Ventilation started','Ventilation started','Tracheostomy','Tracheostomy','Outcome'),
               `Event 2` = c('Outcome','Tracheostomy','Outcome','Death','Death'),
               `Assummed first` = c('Ventilation started','Ventilation started','Tracheostomy','Tracheostomy','Outcome'))
pander(f)
```

We censored patients who did not experience the outcome of interest, for example, because they died. We censored them at the last known date they were in ICU, using the available information on their dates of: ventilation, tracheostomy, speaking valve, first physio appointment, first walking, first meal and out-of-bed exercise.

We used R version 3.5.2 for all analyses. The R code is available here: https://github.com/agbarnett/Tracheostomy.

# Demographics

## Table of categorical variables

```{r demo}
# table
cat.tab = tabular( (Heading('Gender')*Gender + 
                    Heading('Primary diagnosis')*diagnosis +
                    Heading('ICU survival')*ICUsurv +
                    Heading('Long-term dead or alive')*dead
                      ) + 1~((n=1) +  Heading('%')*Percent('col')*Format(digits=0)), data=data) 
pander(cat.tab)
```

## Table of continuous variables

```{r mega.table.cont}
cont.tab = tabular(Heading('Age (years)')*Age 
                   + Heading('Weight (kg)')*Weight
                   + Heading('BMI (kg/m2)')*BMI 
                   + APACHE 
                   + Heading('Time from ventilation to tracheostomy (days)')*time.to.t
                   + Heading('SOFA 1')*SOFA1
                   + Heading('SOFA 2')*SOFA2
                   + Heading('SOFA 3')*SOFA3
                   + Heading('SOFA 4')*SOFA4
                   + Heading('Total ventilation days')*ventdays 
                   + Heading("Best mobility score when first out of bed exercise")*mobscorebed 
                   + Heading("Best mobility score on discharge")*mobscoredis
                   + Heading('Long term survival (years)')*surv.time.years
                   ~ (Missing + N + Mean*Format(digits=2) + SD*Format(digits=2)+ Min*Format(digits=0)+ Max*Format(digits=0)), data=data)
pander(cont.tab)
```

Some variables have a relatively high number of missing values, in particular BMI.

## Histogram of time from ventilation to tracheostomy

```{r dist}
hplot = ggplot(data=data, aes(x=time.to.t)) +
  geom_histogram(fill='skyblue')+
  xlab('Time from ventilation to tracheostomy (days)')+
  ylab('Count')+
  theme_bw()
hplot
```

There is a positive skew in the time between ventilation and tracheostomy.

# Events in ICU

Here we examine the timing of events in the ICU and test whether having a tracheostomy was associated with shorter times for some events.

```{r include=F}
# some basic set up
my.curvlab = c("Tracheostomy by day d", "No tracheostomy by day d")
my.xlab = 'Days since ventilation, d'
```

## Time between tracheostomy and food commenced
 
```{r food, fig.width=6, fig.height=5, dpi=300}
e = extra.time(indata = data, outcome = 'food', n.boot=100)
my.ylab = 'Days to food commenced'
plot(e$cLOS, curvlab=my.curvlab, col.e=cbPalette[2:3], xlab=my.xlab, ylab.e=my.ylab)
pander(select(e$stats, -outcome) %>% rename('Number censored' = 'n.censored'), digits = c(0,2,2,2))
```

The bottom panel shows the days since ventilation on the x-axis and the remaining expected days until food commenced on the y-axis. We can see that at every day, those who already had a tracheostomy had a shorter remaining time until they commenced food compared with those without a tracheostomy. The top panel shows the weights, which are the times at which patients received the tracheostomy.

The table shows the estimated change in length of stay and associated 95% confidence interval. There was a reduction in the time to commencing food for patients with a tracheostomy, with an average reduction of `r roundz(e$stats$mean,1)` days compared with patients without a tracheostomy. The 95% confidence interval shows that this reduction is strongly statistically significant as the interval does not include zero.


## Time between tracheostomy and speaking valve

```{r speaking, fig.width=6, fig.height=5, dpi=300}
e = extra.time(indata = data, outcome = 'SVDate', n.boot=100)
my.ylab = 'Days to speaking valve'
plot(e$cLOS, curvlab=my.curvlab, col.e=cbPalette[2:3], xlab=my.xlab, ylab.e=my.ylab)
pander(select(e$stats, -outcome) %>% rename('Number censored' = 'n.censored'), digits = c(0,2,2,2))
```

## Time between tracheostomy and first physio 

```{r physio, fig.width=6, fig.height=5, dpi=300}
e = extra.time(indata = data, outcome = 'physio', n.boot=100)
my.ylab = 'Days to first physio'
plot(e$cLOS, curvlab=my.curvlab, col.e=cbPalette[2:3], xlab=my.xlab, ylab.e=my.ylab)
pander(select(e$stats, -outcome) %>% rename('Number censored' = 'n.censored'), digits = c(0,2,2,2))
```

## Time between tracheostomy and first out of bed exercise

```{r exercise, fig.width=6, fig.height=5, dpi=300}
e = extra.time(indata = data, outcome = 'exercise', n.boot=100)
my.ylab = 'Days to first exercise'
plot(e$cLOS, curvlab=my.curvlab, col.e=cbPalette[2:3], xlab=my.xlab, ylab.e=my.ylab)
pander(select(e$stats, -outcome) %>% rename('Number censored' = 'n.censored'), digits = c(0,2,2,2))
```


# Short- and long-term survival

We examine two time-frames using survival analysis: 1) during ICU (short-term), and 2) for those that were discharged alive from ICU (long-term).  

Long-term survival was monitored from the date of tracheostomy up to `r format(as.Date(last.alive.date, origin='1899-12-30'), '%d-%b-%Y')`. Patients that returned to ICU and received another tracheostomy had their first survival time censored at the time of the second tracheostomy. 

We split the Kaplan-Meier survival curves by early vs late tracheostomy using a split of: 

* up to 7 days as early 
* 7 or more days as late


```{r groups, include=FALSE}
# new cohort of just survivors from ICU:
survived = filter(data, ICUsurv == 'Yes')
# create new variable that measures ICU survival time, from tracheostomy date to dead or last ICU date
data = mutate(data, 
              surv.time.ICU = ifelse(ICUsurv=="No", DeathDate - tdate, max.ICU.date - tdate))
```

## Kaplan-Meier plot of survival time after tracheostomy split by early vs late tracheostomy during ICU stay

```{r km, fig.width=7, fig.height=6}
km = survfit(Surv(surv.time.ICU, ICUsurv=="No") ~ early.late, data=data) # whole cohort
res.short = ggsurvplot(km, data = data, risk.table="nrisk_cumevents")
res.short$plot = res.short$plot + xlab('Time (days)')
print(res.short)
```

The vertical notches on the survival curves are censored patients, which will be those patients who were discharged from ICU and hence no longer at risk.
The two survival curves are close

### Long-term survival plot split by early vs late tracheostomy for those who were discharged alive from ICU

```{r km.split, fig.width=7, fig.height=6}
km = survfit(Surv(surv.time.years, dead=="Dead") ~ early.late, data=survived) # cohort of just those who survived ICU
res = ggsurvplot(km, data = data, risk.table="nrisk_cumevents")
res$plot = res$plot +  xlab('Time (years)')+
  coord_cartesian(ylim=c(0.6, 1)) # reduce white space on y-axes
print(res)
# log-rank test
ltest = survdiff(Surv(surv.time.years, dead=="Dead") ~ early.late, data=survived)
ltest$pval = format.pval(1 - pchisq(ltest$chisq, 1), eps=0.001, digits=3)
```

The vertical notches on the survival curves are censored patients, which are those patients still alive at the date their records were checked against the national death index.
The survival curves do appear somewhat different, with a better long-term survival for those with a later tracheostomy. However, the log-rank test does not show a statistical difference (Chi-squared statistic = `r roundz(ltest$chisq,3)`, P-value = `r ltest$pval`).
Note the y-axis does not start at a survival of zero.

### Regression model for early vs late tracheostomy and long-term survival for those who were discharged alive from ICU

Here we use a multiple regression model to estimate the effect of early vs late tracheostomy on long-term survival. We adjust for patient age and gender. We use a Cox survival model and so the estimates are hazard ratios together with 95% confidence intervals (CIs).

```{r cox}
cox = coxph(Surv(surv.time.years, dead=="Dead") ~ time.to.t + I((Age-50)/10) + Gender, data=survived)
to.table = tidy(cox) %>%
  dplyr::select('term','estimate','conf.low','conf.high','p.value') %>%
  mutate(estimate = exp(estimate),
         conf.low = exp(conf.low),
         conf.high = exp(conf.high),
         p.value = format.pval(p.value, eps=0.001, digits=3)) %>%
  rename('Variable' = 'term',
         `Hazard Ratio` = 'estimate',
         `Lower CI` = 'conf.low',
         `Upper CI` = 'conf.high',)
# rename variables
to.table$Variable[grep('time.to.t', to.table$Variable)] = 'Time to tracheostomy (+1 day)'
to.table$Variable[grep('Age', to.table$Variable)] = 'Age (+10 years)'
to.table$Variable[grep('Gender', to.table$Variable)] = 'Gender = Male'
# table
pander(to.table, digits=c(0,3,3,3,3))
prop.test = cox.zph(cox)
## Looks like time.to.t is non-proportional and should be split after first year (based on graph above)
```

Each 10 year increase in age greatly increases the hazard of death. There was no strong association between tracheostomy timing and the long-term risk of death.

# References

* Arthur Allignol, Martin Schumacher, Jan Beyersmann (2011). Empirical Transition Matrix of Multi-State
  Models: The etm Package. Journal of Statistical Software, 38(4), 1-15. URL
  http://www.jstatsoft.org/v38/i04/.
  
* Jan Beyersmann, Arthur Allignol and Martin Schumacher (2012). Competing Risks and Multistate Models with R. Springer. DOI: 10.1007/978-1-4614-2035-4

* R Core Team (2018). R: A language and environment for statistical computing. R Foundation for Statistical
  Computing, Vienna, Austria. URL https://www.R-project.org/.